---
title: "Swing Clusters Within Localized Strike-Zone Pitch Contexts"
output: 
  posterdown::posterdown_html
---

```{r setup, echo=FALSE, warning=FALSE, message=FALSE}
library(dplyr)
library(readr)
library(ggplot2)
library(knitr)
library(scales)
library(stringr)

analysis_dir <- "output/localized_angles/bat_speed_independent"
validation_dir <- "output/phase_5_cluster_validation/localized_angles/bat_speed_independent"
player_dir <- "output/player_profiles/localized_angles/bat_speed_independent"

metadata <- readRDS(file.path(analysis_dir, "analysis_metadata.rds"))
bucket_counts <- read_csv(file.path(analysis_dir, "bucket_counts.csv"), show_col_types = FALSE)
overall_tests <- read_csv(file.path(analysis_dir, "significance_overall.csv"), show_col_types = FALSE)
pairwise_tests <- read_csv(file.path(analysis_dir, "significance_pairwise.csv"), show_col_types = FALSE)
cluster_summary <- read_csv(file.path(analysis_dir, "cluster_summary.csv"), show_col_types = FALSE)
validation_summary <- read_csv(file.path(validation_dir, "summary.csv"), show_col_types = FALSE)
player_overview <- read_csv(file.path(player_dir, "player_overview.csv"), show_col_types = FALSE)

cluster_count_tbl <- cluster_summary %>%
  distinct(bucket_id, cluster_id) %>%
  count(bucket_id, name = "selected_clusters") %>%
  count(selected_clusters, name = "buckets") %>%
  arrange(selected_clusters)

top_contact <- overall_tests %>%
  filter(outcome == "contact_flag") %>%
  arrange(desc(max_cluster_gap)) %>%
  slice_head(n = 6) %>%
  mutate(bucket_label = factor(bucket_label, levels = rev(bucket_label)))

launch_angle_gap <- cluster_summary %>%
  group_by(bucket_label) %>%
  summarise(
    min_launch_angle = min(launch_angle_on_contact, na.rm = TRUE),
    max_launch_angle = max(launch_angle_on_contact, na.rm = TRUE),
    launch_angle_gap = max_launch_angle - min_launch_angle,
    .groups = "drop"
  ) %>%
  arrange(desc(launch_angle_gap)) %>%
  slice_head(n = 8) %>%
  mutate(
    bucket_label = str_wrap(bucket_label, width = 26),
    bucket_label = factor(bucket_label, levels = rev(bucket_label))
  )

sig_summary <- tibble(
  outcome = c("Contact", "In play", "Launch angle", "xwOBA on contact"),
  overall_sig = c(
    sum(overall_tests$outcome == "contact_flag" & overall_tests$p_value < 0.05, na.rm = TRUE),
    sum(overall_tests$outcome == "in_play_flag" & overall_tests$p_value < 0.05, na.rm = TRUE),
    sum(overall_tests$outcome == "launch_angle" & overall_tests$p_value < 0.05, na.rm = TRUE),
    sum(overall_tests$outcome == "xwoba_on_contact" & overall_tests$p_value < 0.05, na.rm = TRUE)
  ),
  total_buckets = c(81, 81, 81, 81),
  pairwise_sig = c(
    sum(pairwise_tests$outcome == "contact_flag" & pairwise_tests$adjusted_p_value < 0.05, na.rm = TRUE),
    sum(pairwise_tests$outcome == "in_play_flag" & pairwise_tests$adjusted_p_value < 0.05, na.rm = TRUE),
    sum(pairwise_tests$outcome == "launch_angle" & pairwise_tests$adjusted_p_value < 0.05, na.rm = TRUE),
    sum(pairwise_tests$outcome == "xwoba_on_contact" & pairwise_tests$adjusted_p_value < 0.05, na.rm = TRUE)
  ),
  total_pairwise = c(
    sum(pairwise_tests$outcome == "contact_flag"),
    sum(pairwise_tests$outcome == "in_play_flag"),
    sum(pairwise_tests$outcome == "launch_angle"),
    sum(pairwise_tests$outcome == "xwoba_on_contact")
  )
)

player_tbl <- player_overview %>%
  filter(tracked_swings >= 150) %>%
  arrange(desc(contact_vs_bucket_context)) %>%
  slice_head(n = 5) %>%
  transmute(
    Player = player_label,
    `Tracked swings` = tracked_swings,
    `Contact vs context` = round(contact_vs_bucket_context, 3)
)
```

# Main Result: Similar Pitches Still Produced Multiple Swing Types

Holding pitch context constant did not collapse swings into one common solution. Most of the 81 localized pitch buckets supported four swing clusters, and some supported as many as seven.

```{r cluster-count-plot, echo=FALSE, fig.height=3.5}
ggplot(cluster_count_tbl, aes(x = factor(selected_clusters), y = buckets)) +
  geom_col(fill = "#1f5aa6", width = 0.72) +
  geom_text(aes(label = buckets), vjust = -0.3, size = 4) +
  labs(
    x = "Clusters selected within bucket",
    y = "Number of buckets",
    title = "Most buckets supported four swing archetypes"
  ) +
  scale_y_continuous(expand = expansion(mult = c(0, 0.12))) +
  theme_minimal(base_size = 14)
```

# Contact Separation Was the Strongest Outcome Signal

The largest between-cluster differences appeared in contact-related outcomes, especially in lower-zone pitch buckets. In the strongest buckets, the best and worst contact clusters were separated by more than 60 percentage points.

```{r contact-effects-plot, echo=FALSE, fig.height=5}
ggplot(top_contact, aes(x = max_cluster_gap, y = bucket_label)) +
  geom_col(fill = "#c44e52", width = 0.72) +
  geom_text(aes(label = percent(max_cluster_gap, accuracy = 1)), hjust = -0.08, size = 3.6) +
  scale_x_continuous(labels = percent_format(accuracy = 1), expand = expansion(mult = c(0, 0.14))) +
  labs(
    x = "Best-to-worst cluster gap in contact rate",
    y = NULL,
    title = "Largest contact-rate gaps were concentrated low in the zone"
  ) +
  theme_minimal(base_size = 13)
```

# Contact Shape Also Changed After Contact

Cluster differences were not limited to making contact. Some buckets showed large launch-angle gaps across clusters, meaning similar pitch contexts still produced trajectories ranging from ground balls to high fly balls.

```{r launch-angle-gap-plot, echo=FALSE, fig.height=5}
ggplot(launch_angle_gap, aes(y = bucket_label)) +
  geom_segment(aes(x = min_launch_angle, xend = max_launch_angle, yend = bucket_label), linewidth = 1.1, color = "#8ab1e3") +
  geom_vline(xintercept = 0, color = "#b7c4cf", linetype = "dashed") +
  geom_point(aes(x = min_launch_angle), size = 2.7, color = "#d1495b") +
  geom_point(aes(x = max_launch_angle), size = 2.7, color = "#2a9d8f") +
  geom_text(aes(x = max_launch_angle + 2, label = paste0(round(launch_angle_gap, 1), "°")), hjust = -0.1, size = 3.4) +
  labs(
    x = "Average launch angle on contact",
    y = NULL,
    title = "Largest launch-angle gaps within the same pitch bucket"
  ) +
  theme_minimal(base_size = 13) +
  theme(panel.grid.major.y = element_blank())
```

# Widespread Significance, Different Practical Strength

```{r significance-table, echo=FALSE}
kable(
  sig_summary %>%
    mutate(
      `Overall significant` = paste0(overall_sig, "/", total_buckets),
      `Pairwise significant` = paste0(pairwise_sig, "/", total_pairwise)
    ) %>%
    select(
      Outcome = outcome,
      `Overall significant`,
      `Pairwise significant`
    ),
  caption = "Between-cluster outcome differences"
)
```

xwOBA on contact also differed across clusters, but its overall effect sizes were usually smaller and more distributed than contact rate or in-play rate. The clearest signal was swing shape separating contact and fair-contact success.

# Stable Clusters, Not Perfectly Separate Categories

The cluster structure was moderately separated but highly stable. That supports interpretation as repeatable swing archetypes, not perfectly isolated natural categories.

```{r validation-box, echo=FALSE}
validation_tbl <- tibble(
  Metric = c("Mean silhouette", "Median silhouette", "Mean stability ARI", "Median stability ARI"),
  Value = c(
    round(validation_summary$mean_selected_silhouette[[1]], 3),
    round(validation_summary$median_selected_silhouette[[1]], 3),
    round(validation_summary$mean_stability_ari[[1]], 3),
    round(validation_summary$median_stability_ari[[1]], 3)
  )
)

kable(validation_tbl, caption = "Cluster validation summary")
```

# Player-Level Signal

The same framework scaled to individual hitters. Several recognizable contact-oriented players stood out as better than expected after adjusting for bucket context.

```{r player-table, echo=FALSE}
kable(player_tbl, caption = "Top context-adjusted contact performers")
```

# Final Takeaways

- Similar in-zone pitches still produced multiple swing archetypes.
- Contact rate and in-play rate were the clearest outcome separators.
- Launch angle showed that clusters also differed in contact shape.
- xwOBA on contact differed too, but less cleanly than contact outcomes.
- The clusters were stable enough to support league-level and player-level interpretation.
